---
title: "Regression Analysis Project"
author: "Jonathan Bennett"
output: html_document
---

##Table of Contents
REPORT  
1. Executive Summary  
2. Coefficient Interpretataion
  
APPENDIX  
3. Exploratory Analysis  
4. Model Selection  
5. Residual Plot  
6. Model Comparison    

##1. Executive Summary
This analysis addresses the relationship between a group of variables and the resulting miles per
gallon for a subset of cars from the mtcars dataset.  Specificically the analysis determines whether
an automatic or manual transmission achieves better gas mileage, and attempts to quantify the
difference.  For details on the dataset, please visit:  
https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html  
  
Findings:
Using the AIC approach, an automatic transmission decreased gas mileage by 0.07 MPG.  Appropriate
input variables were wt, cyl, hp, and, of course, am.  However, the standard error for the am
is 1.4, meaning that a 95% confidence interval spans -2.84 to 2.69.  Therefore the finding that
an automatic transmission is less efficient can't be stated with a 95% guarantee.  Rather, the
transmission type is inconclusive as a predictor of MPG.
  
This analysis employs a rigorous quantitative and qualatitive approach to model selection.  This
investigates five model approaches:  
1) Use all regression variables (cyl, disp, hp, drat, wt, qsec, am, gear, vs, carb)  
2) Akaike Information Criterion (AIC-http://en.wikipedia.org/wiki/Akaike_information_criterion)  
3) Bayesian Information Criterion (BIC-http://en.wikipedia.org/wiki/Bayesian_information_criterion)  
4) Only use am as a predictor  
  
The AIC approach made the most sense because it optimizes fit without overfitting.  Other approaches
minimized rsquared values, but introducing variables uncorrelated to the output will increase 
rsquared values so minimizing rsquared was not a determining factor.  This  analysis performed some
data maniuplation prior to fitting:  
1) Converted appropriate inputs to factor type (am, cyl, gear, car, vs).  This made sense because
the data were not numeric in nature.  Arguably gear, carb and cyl may be, but am and vs are most
certainly not.  
2) Removed training outliers that were outside of the 95th percentile.  Given the few data points,
outliers were likely to skew results so the outliers were deemed as detrimental to a good fit for
purposes of this investigation.  
  
For purposes of brevity, much of the code in this Rmd file has been hidden as the assignment
mandated a length of less than seven pages total.

##2. Coefficient Interpretation
The regression equation to obtain the mpg is:
mpg = 35.92996 - 3.70181 * wt - 0.02670 * hp - 1.35179 (if 6 cylinders) - 1.16743 (if 8 cylinders)
- 0.07253 (if automatic transmission)  
We start at 35.92996 MPG (intercept) before adding other elements.  It makes sense the we lose
appx 3.70 MPG for each unit of weight, and 1.35 for each unit of hp.  A four cylinder car has no MPG
adder, but both 6 and 8 cylinders reduce gas mileage.  The regression equation indicates 8 cylinder
cars are more efficient (less or a reduction) than 6 cylinder cars, although this might not folow
intuition.  More data or analysis is warranted on this variable.  

```{r globals, echo=FALSE}
require(ggplot2)
require(reshape)
require(leaps)

confidence = 1.96 # Exclude datapoints that don't fall within the 95% confidence interval

filteroutlier <- function(df,zvalue=3){
    #receives a numeric dataframe
    #returns a boolean dataframe indicating whether each element is between plus or minus z sds
    #default is three standard deviations if parameter is not provided
    output <- (df <= (zvalue*sd(df)+mean(df))) & ((df >= (-zvalue*sd(df)+mean(df))))
}
```

##3. Exploratory Analysis
```{r dataexplore}
d <- melt(mtcars)
ggplot(d,aes(x = value)) +
    facet_wrap(~variable,scales = "free_x") +
    geom_histogram()

# d <- plot(mtcars$mpg) #not generated for purposes of brevity

#look at outliers and remove all datapoint outside of +- two standard deviations
boxplot(mtcars)
outliers <- !((filteroutlier(mtcars$mpg,confidence)) & (filteroutlier(mtcars$disp,confidence)) &
                  (filteroutlier(mtcars$hp,confidence)) & (filteroutlier(mtcars$drat,confidence)) &
                  (filteroutlier(mtcars$wt,confidence)) & (filteroutlier(mtcars$qsec,confidence)))
mtcars2 <- mtcars[!outliers,]

```
##4. Model Selection
```{r modelselection, echo = FALSE}
mtcars3 <- mtcars2
mtcars3$am <- as.factor(mtcars3$am)
mtcars3$cyl <- as.factor(mtcars3$cyl)
mtcars3$gear <- as.factor(mtcars3$gear)
mtcars3$carb <- as.factor(mtcars3$carb)
mtcars3$vs <- as.factor(mtcars3$vs)

require(leaps)
leaps <- regsubsets(mpg ~ cyl + disp + hp + drat + wt + qsec + vs + am + gear + carb,
                    data = mtcars3, nbest = 3)

#plot(leaps, scale ="bic")

null <- lm(mpg ~ 1, data = mtcars3)
full <- lm(mpg ~ cyl + disp + hp + drat + wt + qsec + am + gear + vs + carb, data = mtcars3)
step(null, scope=list(lower=null, upper=full),direction="both")

```

```{r regression, echo = TRUE}
#Best AIC with transmission variable approach
fit_aic <- lm(mpg ~ wt + cyl + hp + am, data = mtcars3)
#summary(fit_aic)

#BIC approach
fit_bic <- lm(mpg ~ wt + qsec + am, data = mtcars3)

#all variables
fit_all <- lm(mpg ~ cyl + disp + hp + drat + wt + qsec + am + gear + vs + carb, data = mtcars3)

#only am
fit_onlyam <- lm(mpg ~ am, data = mtcars3)
```
```{r results, echo=FALSE}
approach <- c("All_variables","AIC","BIC","only_am")
automatic_mpgadder <- c(coef(fit_all)[names(coef(fit_all))=="am1"],
                        coef(fit_aic)[names(coef(fit_aic))=="am1"],
                        coef(fit_bic)[names(coef(fit_bic))=="am1"],
                        coef(fit_onlyam)[names(coef(fit_onlyam))=="am1"])
rsquared <- c(summary(fit_all)$r.squared,
             summary(fit_aic)$r.squared,
             summary(fit_bic)$r.squared,
             summary(fit_onlyam)$r.squared)
degreesf <- c(length(names(coef(fit_all))),
              length(names(coef(fit_aic))),
              length(names(coef(fit_bic))),
              length(names(coef(fit_onlyam))))
results <- data.frame(approach,automatic_mpgadder,rsquared, degreesf)
```

##5. Residual Plot
```{r plotfit, echo=FALSE}
plot(fit_aic$resid)
```

##6. Model Comparisons
```{r comparisons, echo=FALSE}
results
```