---
title: "Investigating Tooth Growth by supplement"
author: "Jonathan Bennett"
date: "Saturday, March 21, 2015"
output: html_document
---
##Exploratory Data

```{r exploratory}
    coplot(len ~ dose | supp, data = ToothGrowth, panel = panel.smooth,
           xlab = "ToothGrowth data: length vs dose, given type of supplement")
    oj<-ToothGrowth[ToothGrowth[,2]=="OJ",]
    vc<-ToothGrowth[ToothGrowth[,2]=="VC",]
    hist(oj$len)
    hist(vc$len)    
    mean(oj$len)
    sd(oj$len)
    length(oj$len)
    mean(vc$len)
    sd(vc$len)
    length(vc$len)
```
##Confidence Intervals
###The assumption here is that it's normally distributed.
```{r confidenceinterval1}
    ci <- .975
    ojci <- mean(oj$len)+c(-1,1)*qnorm(ci)*sd(oj$len)/sqrt(length(oj$len))
    vcci <- mean(vc$len)+c(-1,1)*qnorm(ci)*sd(vc$len)/sqrt(length(vc$len))
    confidencelevel <- (2*ci-1)*100
```
####The `r confidencelevel`% for OJ is between `r ojci[1]` and `r ojci[2]`.
####The `r confidencelevel`% for Vitamin C is between `r vcci[1]` and `r vcci[2]`.
```{r confidenceinterval2}
    ci <- .95
    ojci <- mean(oj$len)+c(-1,1)*qnorm(ci)*sd(oj$len)/sqrt(length(oj$len))
    vcci <- mean(vc$len)+c(-1,1)*qnorm(ci)*sd(vc$len)/sqrt(length(vc$len))
    confidencelevel <- (2*ci-1)*100
```
####The `r confidencelevel`% for OJ is between `r ojci[1]` and `r ojci[2]`.
####The `r confidencelevel`% for Vitamin C is between `r vcci[1]` and `r vcci[2]`.  


###The assumption here is that it's a poisson distribution.
```{r confidenceinterval3}
    ci <- .95
    ojci <- mean(oj$len)+c(-1,1)*pnorm(ci)*sd(oj$len)/sqrt(length(oj$len))
    vcci <- mean(vc$len)+c(-1,1)*pnorm(ci)*sd(vc$len)/sqrt(length(vc$len))
    confidencelevel <- (2*ci-1)*100
```
####The `r confidencelevel`% for OJ is between `r ojci[1]` and `r ojci[2]`.
####The `r confidencelevel`% for Vitamin C is between `r vcci[1]` and `r vcci[2]`.

```{r confidenceinterval4}
    ci <- .95
    ojci <- mean(oj$len)+c(-1,1)*pnorm(ci)*sd(oj$len)/sqrt(length(oj$len))
    vcci <- mean(vc$len)+c(-1,1)*pnorm(ci)*sd(vc$len)/sqrt(length(vc$len))
    confidencelevel <- (2*ci-1)*100
```
####The `r confidencelevel`% for OJ is between `r ojci[1]` and `r ojci[2]`.
####The `r confidencelevel`% for Vitamin C is between `r vcci[1]` and `r vcci[2]`.